Persistence in functional topology and a correction to a theorem of Morse

### 2021-07-29 abstract.tex
Overleaf version to post

During the 1930s, Marston Morse developed a vast generalization of what is commonly known as Morse theory relating the critical points of a semi-continuous functional with the topology of its sublevel sets.
Morse and Tompkins applied this body of work, referred to as functional topology, to prove the Unstable Minimal Surface Theorem in the setting defined by Douglas' solution to Plateau's Problem.
Several concepts introduced by Morse in this context can be seen as early precursors to the theory of persistent homology, which by now has established itself as a popular tool in applied and theoretical mathematics.
In this article, we provide a modern redevelopment of the homological aspects of Morse's functional topology from the perspective of persistence theory.
We adjust several key definitions and prove stronger statements, including a generalized version of Morse inequalities, in order to allow for novel uses of persistence techniques in functional analysis and symplectic geometry.
As an application, we identify and correct a mistake in the proof of the Unstable Minimal Surface Theorem of Morse and Tompkins.

>>>>

During the 1930s, Marston Morse developed a vast generalization of what is commonly known as Morse theory relating the critical points of a semi-continuous functional with the topology of its sublevel sets.
Morse and Tompkins applied this body of work, referred to as functional topology, to prove the Unstable Minimal Surface Theorem in the setting defined by Douglas' solution to Plateau's Problem.
Several concepts introduced by Morse in this context can be seen as early precursors to the theory of persistent homology, which by now has established itself as a popular tool in applied and theoretical mathematics.
In this article, we provide a modern redevelopment of the homological aspects of Morse's functional topology from the perspective of persistence theory.
We adjust several key definitions and prove stronger statements, including a generalized version of the Morse inequalities, in order to allow for novel uses of persistence techniques in functional analysis and symplectic geometry.
As an application, we identify and correct a mistake in the proof of the Unstable Minimal Surface Theorem by Morse and Tompkins.

### 2021-07-29 connectivity.tex
Overleaf version to post

\section{Persistence diagrams from local connectivity} \label{s:connectivity}

We have seen that q-tame functions admit persistence diagrams, which can be used to formulate Morse inequalities.
With q-tameness being a rather abstract condition, we now establish more concrete topological conditions that ensure the q-tameness of a function.
Our definitions are motivated by similar conditions that Morse considered in his work on functional topology.
We will present a historical account in \cref{s:surfaces}.

@@

Concretely, let $\H = (\H_d)_{d \in \Z} \colon \Top \to \Vect$ be a fixed graded homotopy invariant functor, which we call a \emph{homology theory}.
A triple of spaces $X_{1}, X_{2} \subseteq X$ is said to have a \emph{Mayer-Vietoris sequence} for $\H$ if the inclusion-induced maps fit in a long exact sequence
\begin{equation*}
\begin{tikzcd}[column sep = small]
\cdots \arrow[r] & \H_{n}(X_{1} \cap X_{2}) \arrow[r] & \H_{n}(X_{1}) \oplus \H_{n}(X_{2}) \arrow[r] & \H_{n}(X) \arrow[r] & \H_{n-1}(X_{1} \cap X_{2}) \arrow[r] & \cdots \ .
\end{tikzcd}
\end{equation*}
We say that $\H$ has the \emph{open} (resp.\@\ \emph{compact}) \emph{Mayer-Vietories property} if there are natural Mayer-Vietoris sequences for all triples $X_{1}, X_{2} \subseteq X$ with $X = X_1 \cup X_2$ and $X_i \subseteq X$ open (resp.\@\ compact Hausdorff).

For the rest of this section, we will assume that $\H$ is a homology theory that has either the open or the compact Mayer-Vietoris property and for which there is $n_0$ such that $\H_{n}$ is 0 for all $n \leq n_0$.
Note that this includes singular homology with field coefficients, which, like any homology theory in the sense of the Eilenberg-Steenrod axioms \cite[Section I]{Eilenberg.1952}, has the open Mayer-Vietoris property, and it also includes \v{C}ech homology with field coefficients, which has the compact Mayer-Vietoris property.

@@

\begin{defi}
	The sublevel set filtration of a function $f \colon X \to \R$ is said to be \emph{locally homologically small} or $\HLC$ if for any $x \in X$, any neighborhood $V$ of $x$, and any index $t > f(x)$, there is an index $s$ with $f(x) < s < t$ and a neighborhood $U$ of $x$ with $U \subseteq V$ such that the inclusion $f_{\leq s} \cap U \to f_{\leq t} \cap V$ is $\HS$.
\end{defi}

Similarly, we have the following stronger version of this property.

\begin{defi}
	The sublevel set filtration of a function $f \colon X \to \R$ is called \emph{strongly locally homologically small} or \emph{strongly $\HLC$} if for any $x \in X$, any neighborhood $V$ of $x$, and any pair of indices $s,t$ with $f(x) < s < t$ there is a neighborhood $U$ of $x$ with $U \subseteq V$ such that the inclusion $f_{\leq s} \cap U \to f_{\leq t} \cap V$ is $\HS$.
\end{defi}

Clearly, any strongly $\HLC$ sublevel set filtration is also $\HLC$:
while the $\HLC$ property merely requires the existence of an index $s \in (f(x),t)$ satisfying the $\HS$ condition, the \emph{strong} $\HLC$ property requires the $\HS$ condition to hold for any $s \in (f(x),t)$.
If the filtration is induced by a continuous function, the converse also holds as the following theorem shows.

\begin{thm} \label{t:hlc to strong hlc}
	If the sublevel set filtration of a continuous function $f \colon X \to \R$ is $\HLC$, then it is also strongly $\HLC$.
\end{thm}

@@

	To do so, we start by using the $\HLC$ property to choose a neighborhood $U' \subseteq V$ of $x$ and an index $s' \in (f(x),\, t)$ such that the inclusion $f_{\leq s'} \cap U' \to f_{\leq t} \cap V$ is $\HS$.
	Now, we choose $U = f_{< s'} \cap U'$, where $f_{< s'} = f^{-1} (-\infty, s')$.
	Note that this choice of $U$ still defines a neighborhood of $x$ because $f$ is assumed to be continuous, so that $f_{< s'}$ is an open subset of $X$.

@@

As our main result, we will now show that for compact sublevel set filtrations the strong $\HLC$ condition implies \mbox{q-tameness}, and consequently also implies the existence of a persistence diagram which yields persistent Morse inequalities.

\begin{thm} \label{t:strong local connectedness implies q-tameness}
	If the sublevel set filtration of a function $f \colon X \to \R$ is compact and strongly $\HLC$, then it is also q-tame.
\end{thm}

The general proof strategy is inspired by the proof of a result attributed to Wilder as presented by Bredon \cite[Section II.17]{Bredon.1997}.
We collect the main ideas in several lemmas.

@@

We want to use \cref{l:neighborhood third} on the domain of the function whose sublevel set filtration we consider. However, we do not want to assume the domain to be locally compact for \cref{t:strong local connectedness implies q-tameness}. To circumvent this, we will work in one of the sublevel sets, which are assumed to be compact Hausdorff and hence locally compact. This requires the use of a slight weakening of the strong $\HLC$ condition.

\begin{defi}
	For $u \in \R$, the sublevel set filtration of a function $f \colon X \to \R$ is called \emph{strongly $\HLC$ below $u$} if for any $x \in X$, any neighborhood $V$ of $x$, and any pair of indices $s,t$ with $f(x) < s < t < u$ there is a neighborhood $U$ of $x$ with $U \subseteq V$ such that the inclusion $f_{\leq s} \cap U \to f_{\leq t} \cap V$ is $\HS$.
\end{defi}

\begin{lem} \label{l:restriction of LHS filtration}
	Let $f \colon X \to \R$ be a function whose sublevel set filtration is strongly $\HLC$.
	Fix $u \in \R$ and let $g \colon Y \to \R$ be the restriction of $f$ to the sublevel set $Y = f_{\leq u}$.
	Then, the sublevel set filtration defined by $g$ is strongly $\HLC$ below $u$.
\end{lem}

@@

	Since $Y \subseteq X$ carries the subspace topology, we may choose a neighborhood $V'$ of $x$ in $X$ such that $V = V' \cap Y$. 
	The sublevel set filtration of $f$ is assumed to be strongly $\HLC$, so there is a neighborhood $U' \subseteq V'$ of $x$ in $X$ such that the inclusion $f_{\leq s} \cap U' \to f_{\leq t} \cap V'$ is $\HS$.
	We set $U = U' \cap Y$, which defines a neighborhood of $x$ in $Y$.

@@

\begin{lem} \label{l:key lemma for q-tameness}
	Let $f \colon X \to \R$ be a function on a locally compact space $X$ whose sublevel set filtration is compact and strongly $\HLC$ below $u \in \R$, and consider subsets $C \subseteq L \subseteq X$ with $C$ compact and $L$ open.
	For any $s < t < u$ the inclusion $C \cap f_{\leq s} \to L \cap f_{\leq t}$ is $\HS$.
\end{lem}

@@

	Next, we will show using the strong $\HLC$ property that $\Sigma_{s, t}$ has the first required property, i.e., that any point $x \in L \cap f_{\leq s}$ has a neighborhood in $\Sigma_{s, t}$.
	Choose an arbitrary $s' \in (s,\, t)$.
	Since the sublevel set filtration of $f$ is strongly $\HLC$ below $u$ and we have $s' < t < u$, there is an open neighborhood $U_x \subseteq L$ such that the inclusion
	$U_x \cap f_{\leq s'} \to L \cap f_{\leq t}$
	is $\HS$, so in particular $\HS_n$.
	By local compactness of $X$ we can choose a compact neighborhood $K_x$ of $x$ contained in $U_x$.
	Now $V_x = \interior (K_x)$ is a neighborhood of $x$ with $V_x \in \Sigma_{s,t}$.

@@

We can now complete the proof of the claim stating that for compact sublevel set filtrations, strong $\HLC$ implies q-tameness.

\begin{proof}[Proof of \cref{t:strong local connectedness implies q-tameness}]
	By definition, the sublevel set filtration of $f$ is q-tame if and only if the inclusion $f_{\leq s} \to f_{\leq t}$ is $\HS$ for all pairs $s < t$. 
	Choose $u \in \R$ with $u > t$ and let $g \colon Y \to \R$ be the restriction of $f$ to the sublevel set $Y=f_{\leq u}$. 
	Since we assume $f$ to induce a strongly $\HLC$ sublevel set filtration, by \cref{l:restriction of LHS filtration} the sublevel set filtration of $g$ is strongly $\HLC$ below $u$. 
	Clearly, the sublevel set filtration of $g$ is also compact, and its domain $Y$ is locally compact being a compact Hausdorff space by assumption.
	Thus, we can apply \cref{l:key lemma for q-tameness} to the filtration $g_{\leq \bullet}$ with $C = f_{\leq s}$ and $L = Y$ to obtain that the inclusion $f_{\leq s} = C \cap g_{\leq s} \to L \cap g_{\leq t} = f_{\leq t}$ is $\HS$.
\end{proof}

Combining \cref{t:hlc to strong hlc} with \cref{t:strong local connectedness implies q-tameness} also yields the following result for continuous functions.

\begin{cor} \label{c:q-tameness for continuous functions}
	If the sublevel set filtration of a continuous function $f \colon X \to \R$ is compact and $\HLC$, then it is also q-tame.
\end{cor}

>>>>

% \section{Sufficient conditions for the existence of persistence diagrams} \label{s:connectivity}

\section{Existence of persistence diagrams} \label{s:connectivity}

We have seen that q-tame functions admit persistence diagrams, which can be used to formulate Morse inequalities.
With q-tameness being a rather abstract condition, we now establish concrete topological requirements that ensure the q-tameness of a function.
Our definitions are motivated by similar conditions that Morse considered in his work on functional topology.
We will present a historical account in \cref{s:surfaces}.

@@

Concretely, let $\H = (\H_d)_{d \in \Z} \colon \Top \to \Vect$ be a fixed graded homotopy invariant functor, which we call a \emph{homology theory}.
A triple of spaces $X_{1}, X_{2} \subseteq X$ is said to have a \emph{Mayer-Vietoris sequence} for $\H$ if the inclusion-induced maps assemble to a long exact sequence
\begin{equation*}
\begin{tikzcd}[column sep = small]
\cdots \arrow[r] & \H_{n}(X_{1} \cap X_{2}) \arrow[r] & \H_{n}(X_{1}) \oplus \H_{n}(X_{2}) \arrow[r] & \H_{n}(X) \arrow[r] & \H_{n-1}(X_{1} \cap X_{2}) \arrow[r] & \cdots \ .
\end{tikzcd}
\end{equation*}
We say that $\H$ has the \emph{open} (resp.\@\ \emph{compact}) \emph{Mayer-Vietories property} if there are natural Mayer-Vietoris sequences for all triples $X_{1}, X_{2} \subseteq X$ with $X = X_1 \cup X_2$ and $X_i \subseteq X$ open (resp.\@\ compact Hausdorff).

For the rest of this section, we will assume that $\H$ is a homology theory that has either the open or the compact Mayer-Vietoris property and for which there is $n_0$ such that $\H_{n}$ is 0 for all $n \leq n_0$.
Note that this includes singular homology with field coefficients, which has the open Mayer-Vietoris property (like any homology theory in the sense of the Eilenberg-Steenrod axioms \cite[Section I]{Eilenberg.1952}) , and it also includes \v{C}ech homology with field coefficients, which has the compact Mayer-Vietoris property.

@@

\begin{defi}
	The sublevel set filtration of a function $f \colon X \to \R$ is called \emph{%strongly
	locally homologically small} or \emph{%strongly
	$\HLC$} if for any $x \in X$, any neighborhood $V$ of $x$, and any pair of indices $s,t$ with $f(x) < s < t$ there is a neighborhood $U$ of $x$ with $U \subseteq V$ such that the inclusion $f_{\leq s} \cap U \to f_{\leq t} \cap V$ is $\HS$.
\end{defi}

Similarly, we have the following weaker version of this property.

\begin{defi}
	The sublevel set filtration of a function $f \colon X \to \R$ is said to be \emph{weakly locally homologically small} or \emph{weakly $\HLC$} if for any $x \in X$, any neighborhood $V$ of $x$, and any index $t > f(x)$, there is an index $s$ with $f(x) < s < t$ and a neighborhood $U$ of $x$ with $U \subseteq V$ such that the inclusion $f_{\leq s} \cap U \to f_{\leq t} \cap V$ is $\HS$.
\end{defi}

Clearly, any $\HLC$ sublevel set filtration is also weakly $\HLC$:
while the weak $\HLC$ property merely requires the existence of an index $s \in (f(x),t)$ satisfying the $\HS$ condition, the %\emph{strong} 
$\HLC$ property requires the $\HS$ condition to hold for any $s \in (f(x),t)$.
If the filtration is induced by a continuous function, the converse also holds as the following theorem shows.

\begin{thm} \label{t:weak hlc to hlc}
	If the sublevel set filtration of a continuous function $f \colon X \to \R$ is weakly $\HLC$, then it is also %strongly 
	$\HLC$.
\end{thm}

@@

	To do so, we start by using the weak $\HLC$ property to choose a neighborhood $U' \subseteq V$ of $x$ and an index $s' \in (f(x),\, t)$ such that the inclusion $f_{\leq s'} \cap U' \to f_{\leq t} \cap V$ is $\HS$.
	Now, we choose $U = f_{< s'} \cap U'$, where $f_{< s'} = f^{-1} (-\infty, s')$.
	Note that this choice of $U$ still defines a neighborhood of $x$ because $f$ is assumed to be continuous, so that $f_{< s'}$ is an open subset of $X$.

@@

As our main result we will show now that for compact sublevel set filtrations the $\HLC$ condition implies \mbox{q-tameness} and, consequently, also the existence of a persistence diagram.

\begin{thm} \label{t:local connectedness implies q-tameness}
	If the sublevel set filtration of a function $f \colon X \to \R$ is compact and	$\HLC$, then it is also q-tame.
\end{thm}

The general proof strategy is inspired by the proof of a result attributed to Wilder \cite[p.~325]{Wilder.1949}, as presented by Bredon \cite[Section II.17]{Bredon.1997}.
We collect the main ideas in several lemmas.

@@

We want to use \cref{l:neighborhood third} on the domain of the function whose sublevel set filtration we consider. However, we do not want to assume the domain to be locally compact for \cref{t:local connectedness implies q-tameness}. To circumvent this, we will work in one of the sublevel sets, which are assumed to be compact Hausdorff and hence locally compact.
This requires the use of a slight weakening of the $\HLC$ condition.

\begin{defi}
	For $u \in \R$, the sublevel set filtration of a function $f \colon X \to \R$ is called \emph{%strongly
	$\HLC$ below $u$} if for any $x \in X$, any neighborhood $V$ of $x$, and any pair of indices $s,t$ with $f(x) < s < t < u$ there is a neighborhood $U$ of $x$ with $U \subseteq V$ such that the inclusion $f_{\leq s} \cap U \to f_{\leq t} \cap V$ is $\HS$.
\end{defi}

\begin{lem} \label{l:restriction of LHS filtration}
	Let $f \colon X \to \R$ be a function whose sublevel set filtration is %strongly 
	$\HLC$.
	Fix $u \in \R$ and let $g \colon Y \to \R$ be the restriction of $f$ to the sublevel set $Y = f_{\leq u}$.
	Then the sublevel set filtration defined by $g$ is %strongly 
	$\HLC$ below $u$.
\end{lem}

@@

	Since $Y \subseteq X$ carries the subspace topology, we may choose a neighborhood $V'$ of $x$ in $X$ such that $V = V' \cap Y$. 
	The sublevel set filtration of $f$ is assumed to be %strongly 
	$\HLC$, so there is a neighborhood $U' \subseteq V'$ of $x$ in $X$ such that the inclusion $f_{\leq s} \cap U' \to f_{\leq t} \cap V'$ is $\HS$.
	We set $U = U' \cap Y$, which defines a neighborhood of $x$ in $Y$.

@@

\begin{lem} \label{l:key lemma for q-tameness}
	Let $f \colon X \to \R$ be a function on a locally compact space $X$ whose sublevel set filtration is compact and %strongly 
	$\HLC$ below $u \in \R$, and consider subsets $C \subseteq L \subseteq X$ with $C$ compact and $L$ open.
	For any $s < t < u$ the inclusion $C \cap f_{\leq s} \to L \cap f_{\leq t}$ is $\HS$.
\end{lem}

@@

	Next, we will show using the %strong 
	$\HLC$ property that $\Sigma_{s, t}$ has the first required property, i.e., that any point $x \in L \cap f_{\leq s}$ has a neighborhood in $\Sigma_{s, t}$.
	Choose an arbitrary $s' \in (s,\, t)$.
	Since the sublevel set filtration of $f$ is %strongly
	$\HLC$ below $u$ and we have $s' < t < u$, there is an open neighborhood $U_x \subseteq L$ such that the inclusion
	$U_x \cap f_{\leq s'} \to L \cap f_{\leq t}$
	is $\HS$, so in particular $\HS_n$.
	By local compactness of $X$ we can choose a compact neighborhood $K_x$ of $x$ contained in $U_x$.
	Now $V_x = \interior (K_x)$ is a neighborhood of $x$ with $V_x \in \Sigma_{s,t}$.

@@

We can now complete the proof of the claim stating that for compact sublevel set filtrations, %strong 
$\HLC$ implies q-tameness.

\begin{proof}[Proof of \cref{t:local connectedness implies q-tameness}]
	By definition, the sublevel set filtration of $f$ is q-tame if and only if the inclusion $f_{\leq s} \to f_{\leq t}$ is $\HS$ for all pairs $s < t$. 
	Choose $u \in \R$ with $u > t$ and let $g \colon Y \to \R$ be the restriction of $f$ to the sublevel set $Y=f_{\leq u}$. 
	Since we assume $f$ to induce a %strongly 
	$\HLC$ sublevel set filtration, by \cref{l:restriction of LHS filtration} the sublevel set filtration of $g$ is %strongly 
	$\HLC$ below $u$. 
	Clearly, the sublevel set filtration of $g$ is also compact, and its domain $Y$ is locally compact being a compact Hausdorff space by assumption.
	Thus, we can apply \cref{l:key lemma for q-tameness} to the filtration $g_{\leq \bullet}$ with $C = f_{\leq s}$ and $L = Y$ to obtain that the inclusion $f_{\leq s} = C \cap g_{\leq s} \to L \cap g_{\leq t} = f_{\leq t}$ is $\HS$.
\end{proof}

Combining \cref{t:weak hlc to hlc} with \cref{t:local connectedness implies q-tameness} also yields the following result for continuous functions.

\begin{cor} \label{c:q-tameness for continuous functions}
	If the sublevel set filtration of a continuous function $f \colon X \to \R$ is compact and weakly $\HLC$, then it is also q-tame.
\end{cor}
